pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-07 manuscript/greenwell.tex
switched to ggplot2 in SVM example

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd.lstat.rm <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd.lstat.rm)
plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument. To illustrate, the following snippet of code computes the partial dependence of median home value on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

@@

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, .progress = "text",
                 grid.resolution = 101)
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}
lattice::levelplot(y ~ Petal.Width * Petal.Length | Species, data = pd)
\end{example}

>>>>

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument. To illustrate, the following snippet of code computes the partial dependence of median home value on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, "rm", plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

@@

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}. Note that we had to supply the original training data to \code{partial} through the \code{train} option. This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, .progress = "text",
                 grid.resolution = 101)
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
  geom_tile() + 
  # coord_equal() +
  geom_contour(color = "white", alpha = 0.5) + 
  scale_fill_distiller(palette = "Spectral", na.value = "white") + 
  theme_bw() +
  facet_grid(~ Species)
\end{example}
